---
title: "Surface Roughness"
author: "Jared Stabach, Smithsonian Conservation Biology Institute"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calculate Surface Roughness

We assumed that Addax would select inter-dunal depressions. Instead of including the raw SRTM (elevation) data in our models, we created a variable from the raw elevation termed *Surface Roughness*. This variable is the change in elevation range between pixels. That is, it is the difference between the minimum and maximum values of a cell and its eight surrounding neighbors. Pixels with similar neighborhood values have a low change in elevation range (e.g., flat areas), whereas highly variables neighborhoods have a large change in elevation range (e.g., steep slopes) (Wilson, O’Connell, Brown, Guinan, & Grehan, 2007).

Wilson, M.F.J., O'Connell, B., Brown, C., Guinan, J.C., Grehan, A.J., 2007. Multiscale terrain analysis of multibeam bathymetry data for habitat mapping on the continental slope. Marine Geodesy 30: 3-35.

Original data were downloaded from [EarthExplorer](http://earthexplorer.usgs.gov/). Additional information on SRTM can be found [here](https://lta.cr.usgs.gov/SRTMVF). I selected a study area window and the data product in the [EarthExplorer](http://earthexplorer.usgs.gov/) application. The data were then bulk downloaded using the USGS [bda](https://earthexplorer.usgs.gov/bulk/) tool. 

30-m data often have no data areas (i.e., "holes"). These are errors that must be further processed. To fill these areas, I also downloaded an ancillary dataset (the 90-m SRTM gap-filled product (also available on the[EarthExplorer}(http://earthexplorer.usgs.gov/) site). In ArcGIS, I used a conditional statement to fill the no data areas with the 90-meter data (con(IsNull(srtm_30m),srtm_90m,srtm_30m). Data were projected to UTM Zone 32 North, WGS84 datum (30-meter spatial resolution). 

### Load Libraries

```{r Lib, eval=T, warning=F, message=F}
# Remove everything in memory
rm(list=ls())

# Load raster and rgdal library
library(raster)
library(sp)
library(rgdal)
library(proj4)
```

### Set Working Directory

```{r WDir, eval = F}
getwd()
```

### Load Data

This is a dataset subset of the full study area.

```{r Load, eval=T}
# Load the 30-m SRTM dataset
#srtm <- raster(paste0(getwd(),"/Data/srtm_30m"))
srtm <- raster("Data/srtm_30m")

# Check resolution
res(srtm)

# The project of the file is already defined
srtm

# Plot the result
#plot(srtm)
```

Now load the spatial point data

```{r Spatial Data, eval=T}

# Load the point data
Sub <- readOGR(dsn="Data", layer="Clipped_Add")
#plot(Sub,pch=15,cex=0.7,add=TRUE)
# Look at the file
head(Sub)
# Look at the parameters 
Sub

# What if you had a .csv file
# Load your csv, then make spatial
MyData <- read.csv(file="Data/Clipped_Add_csvFile.csv",header=TRUE,sep=",")
head(MyData)
# Make SpatialPointsFile
xy <- MyData[,c(5,6)]
MyData.Spatial <- SpatialPointsDataFrame(coords= xy, data = MyData, 
                                         proj4string = CRS("+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"))

# Look at the file
head(MyData.Spatial)
# Look at the parameters
MyData.Spatial

#plot(MyData.Spatial,pch=10,cex=0.5,col="red",add=TRUE)
```

Plot the results

```{r PlotSRTM, eval=T}
# Plot the result
plot(srtm)
plot(Sub,pch=15,cex=0.7,add=TRUE)
plot(MyData.Spatial,pch=10,cex=0.5,col="red",add=TRUE)
```

### Generate Terrain Variables

```{r Roughness, eval = T}
rough <- terrain(srtm,opt='roughness')
# Other options. Use help(terrain) for more information
#TRI <- terrain(srtm,opt='TRI')
#TPI <- terrain(srtm,opt='TPI')

plot(rough)
#plot(TRI)
#plot(TPI)

# Other variables that could be calculated
#slope <- terrain(srtm,opt='slope', neighbors=8)
#aspect <- terrain(srtm,opt='aspect', neighbors=8)
# Or:
#x <- terrain(srtm, opt=c('slope', 'aspect'), unit='degrees')
#plot(x)
#hill <- hillShade(slope,aspect,angle=45,direction=0)

# Plot the hillshade and overlay the elevation (alpha is the transparency)
#plot(hill,col=grey(0:100/100),legend=FALSE,main='Niger')
#plot(srtm,col=rainbow(25,alpha=0.35), add=TRUE)
```

### Variable Extraction

```{r Extract, eval=T}
# Extract SRTM values at the point locations
# Take the average roughness within a 2.5 km buffer
Sub$ROUGH <- extract(rough,Sub,method='simple',buffer=2500,fun=mean)
#Sub$TRI <- extract(TRI,Sub,method='simple',buffer=2500,fun=mean)
#Sub$TPI <- extract(TPI,Sub,method='simple',buffer=2500,fun=mean)

# How is this different than just extracting at the point?
Sub$RGH_PT <- extract(rough,Sub)

# Look at the data
head(Sub)

# You could do the same with the .csv file you loaded above
#MyData.Spatial$ROUGH <- extract(rough,MyData.Spatial,method='simple',buffer=2500,fun=mean)
#MyData.Spatial$RGH_PT <- extract(rough,MyData.Spatial)

# Look at the Data
#head(MyData.Spatial)
```
